---
title: "Predicting Highly Rated IMDb Movies"
author: "Rylee Wang"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
tidymodels_prefer()
set.seed(123)
```

# Project Overview

## Research Question

**Can movie metadata predict whether an IMDb movie receives an average rating of 7.0 or higher?**

This project uses four sampled IMDb datasets and compares logistic regression with a decision tree. The analysis looks for associations, not causal effects.

# Data

Place these files in the `data/` folder:

- `title.basics.csv`
- `title.akas.csv`
- `title.crew.csv`
- `title.ratings.tsv`

```{r load-data}
basics <- read_csv(
  "data/title.basics.csv",
  show_col_types = FALSE
)

akas <- read_csv(
  "data/title.akas.csv",
  show_col_types = FALSE
)

crew <- read_csv(
  "data/title.crew.csv",
  show_col_types = FALSE
)

ratings <- read_tsv(
  "data/title.ratings.tsv",
  show_col_types = FALSE
)
```

# Data Preparation

```{r rename-identifiers}
basics <- basics %>% rename(tid = 'title identifier')
akas <- akas %>% rename(tid = titleId)
crew <- crew %>% rename(tid = 'title identifier')
ratings <- ratings %>% rename(tid = tconst)
```

## Movie information

```{r prepare-movies}
movies <- basics %>%
  filter(titleType == "movie") %>%
  mutate(
    startYear = na_if(startYear, "\\N"),
    runtimeMinutes = na_if(runtimeMinutes, "\\N"),
    genres = na_if(genres, "\\N"),
    startYear = as.numeric(startYear),
    runtimeMinutes = as.numeric(runtimeMinutes),
    release_decade = floor(startYear / 10) * 10,
    primary_genre = str_extract(genres, "^[^,]+")
  ) %>%
  select(tid, primaryTitle, startYear, release_decade,
         runtimeMinutes, primary_genre)

nrow(movies)
glimpse(movies)
```

## International-title coverage

```{r prepare-akas}
akas_summary <- akas %>%
  mutate(region = na_if(region, "\\N")) %>%
  group_by(tid) %>%
  summarize(
    n_akas = n(),
    n_regions = n_distinct(region, na.rm = TRUE),
    .groups = "drop"
  )

head(akas_summary)
```

## Crew information

```{r prepare-crew}
crew_summary <- crew %>%
  transmute(
    tid,
    has_director = case_when(
      is.na(directors) ~ "No",
      directors == "\\N" ~ "No",
      TRUE ~ "Yes"
    ),
    has_writer = case_when(
      is.na(writers) ~ "No",
      writers == "\\N" ~ "No",
      TRUE ~ "Yes"
    )
  )

head(crew_summary)
```

## Final modeling dataset

```{r build-model-data}
movie_data <- movies %>%
  inner_join(ratings, by = "tid") %>%
  left_join(akas_summary, by = "tid") %>%
  left_join(crew_summary, by = "tid") %>%
  mutate(
    n_akas = replace_na(n_akas, 0L),
    n_regions = replace_na(n_regions, 0L),
    has_director = replace_na(has_director, "No"),
    has_writer = replace_na(has_writer, "No"),
    high_rating = if_else(averageRating >= 7, "High", "Not High"),
    high_rating = factor(high_rating, levels = c("High", "Not High")),
    release_decade = factor(release_decade),
    primary_genre = factor(primary_genre),
    has_director = factor(has_director),
    has_writer = factor(has_writer)
  ) %>%
  filter(!is.na(startYear), !is.na(primary_genre)) %>%
  select(
    tid, primaryTitle, high_rating, averageRating, numVotes,
    startYear, release_decade, runtimeMinutes, primary_genre,
    n_akas, n_regions, has_director, has_writer
  )

glimpse(movie_data)
nrow(movie_data)
count(movie_data, high_rating)
```

# Exploratory Data Analysis

## Rating distribution

```{r rating-distribution}
rating_plot <- ggplot(movie_data, aes(x = averageRating)) +
  geom_histogram(binwidth = 0.5, boundary = 0, fill = "pink", color = "hotpink") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "green") +
  labs(
    title = "Distribution of IMDb Movie Ratings",
    subtitle = "Movies rated 7.0 or higher are classified as High",
    x = "Average IMDb rating",
    y = "Number of movies"
  ) +
  theme_minimal()

rating_plot
ggsave("figures/rating_distribution.png", rating_plot,
       width = 8, height = 5)
```

## Highly rated movies by genre

Because this is a small sample, genres only need at least five movies to appear in the graph.

```{r genre-summary}
genre_summary <- movie_data %>%
  group_by(primary_genre) %>%
  summarize(
    movies = n(),
    high_rating_rate = mean(high_rating == "High"),
    mean_rating = mean(averageRating),
    .groups = "drop"
  ) %>%
  filter(movies >= 5) %>%
  arrange(desc(high_rating_rate))

genre_summary
```

```{r genre-plot}
genre_plot <- ggplot(
  genre_summary,
  aes(x = reorder(primary_genre, high_rating_rate),
      y = high_rating_rate)
) +
  geom_col(fill = "turquoise") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of Highly Rated Movies by Primary Genre",
    subtitle = "Genres represented by at least five movies",
    x = "Primary genre",
    y = "Movies rated 7.0 or higher"
  ) +
  theme_minimal()

genre_plot
ggsave("figures/high_rating_by_genre.png", genre_plot,
       width = 8, height = 5)
```

## International coverage and ratings

```{r coverage-plot}
coverage_plot <- ggplot(movie_data, aes(x = n_regions, y = averageRating)) +
  geom_point(alpha = 0.6, color = "hotpink") +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  labs(
    title = "IMDb Rating and International Title Coverage",
    x = "Number of represented regions",
    y = "Average IMDb rating"
  ) +
  theme_minimal()

coverage_plot
ggsave("figures/coverage_vs_rating.png", coverage_plot,
       width = 8, height = 5)
```

# Train/Test Split

```{r split-data}
set.seed(123)
movie_split <- initial_split(movie_data, prop = 0.80,
                             strata = high_rating)
movie_train <- training(movie_split)
movie_test <- testing(movie_split)
count(movie_train, high_rating)
count(movie_test, high_rating)
```

# Preprocessing

```{r preprocessing-recipe}
movie_recipe <- recipe(
  high_rating ~ runtimeMinutes + release_decade + primary_genre +
    n_akas + n_regions + has_director + has_writer,
  data = movie_train
) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(primary_genre, threshold = 0.03) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

movie_recipe
```

# Model 1: Logistic Regression

```{r logistic-model}
logistic_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

logistic_workflow <- workflow() %>%
  add_recipe(movie_recipe) %>%
  add_model(logistic_model)

logistic_fit <- fit(logistic_workflow, data = movie_train)
logistic_fit
```

# Model 2: Decision Tree

```{r tree-model}
tree_model <- decision_tree(
  tree_depth = 4,
  min_n = 10,
  cost_complexity = 0.01
) %>%
  set_engine("rpart") %>%
  set_mode("classification")

tree_workflow <- workflow() %>%
  add_recipe(movie_recipe) %>%
  add_model(tree_model)

tree_fit <- fit(tree_workflow, data = movie_train)
tree_fit
```

# Predictions

```{r predictions}
make_predictions <- function(fitted_workflow, test_data) {
  test_data %>%
    select(tid, primaryTitle, high_rating) %>%
    bind_cols(predict(fitted_workflow, new_data = test_data,
                      type = "class")) %>%
    bind_cols(predict(fitted_workflow, new_data = test_data,
                      type = "prob"))
}

logistic_predictions <- make_predictions(logistic_fit, movie_test)
tree_predictions <- make_predictions(tree_fit, movie_test)
```

# Model Evaluation

```{r evaluation-metrics}
classification_metrics <- metric_set(
  accuracy, bal_accuracy, precision, recall, f_meas
)

logistic_metrics <- classification_metrics(
  logistic_predictions,
  truth = high_rating,
  estimate = .pred_class,
  event_level = "first"
) %>% mutate(model = "Logistic Regression")

tree_metrics <- classification_metrics(
  tree_predictions,
  truth = high_rating,
  estimate = .pred_class,
  event_level = "first"
) %>% mutate(model = "Decision Tree")

logistic_auc <- roc_auc(
  logistic_predictions,
  truth = high_rating,
  .pred_High,
  event_level = "first"
) %>% mutate(model = "Logistic Regression")

tree_auc <- roc_auc(
  tree_predictions,
  truth = high_rating,
  .pred_High,
  event_level = "first"
) %>% mutate(model = "Decision Tree")

model_results <- bind_rows(
  logistic_metrics, tree_metrics, logistic_auc, tree_auc
) %>% select(model, .metric, .estimate)

model_results
```

## Model comparison

```{r model-comparison}
comparison_plot <- model_results %>%
  filter(.metric %in% c("accuracy", "bal_accuracy", "f_meas", "roc_auc")) %>%
  ggplot(aes(x = model, y = .estimate, fill = .metric)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = round(.estimate, 2)),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  coord_cartesian(ylim = c(0, 1.05)) +
  labs(
    title = "IMDb Rating Prediction: Model Comparison",
    x = NULL,
    y = "Test-set score",
    fill = "Metric"
  ) +
  theme_minimal()

comparison_plot
ggsave("figures/model_comparison.png", comparison_plot,
       width = 9, height = 5)
```

## Confusion matrices

```{r confusion-matrices}
logistic_confusion <- conf_mat(
  logistic_predictions,
  truth = high_rating,
  estimate = .pred_class
)

tree_confusion <- conf_mat(
  tree_predictions,
  truth = high_rating,
  estimate = .pred_class
)

logistic_confusion
tree_confusion
```

```{r logistic-confusion-plot}
autoplot(logistic_confusion, type = "heatmap") +
  labs(title = "Logistic Regression Confusion Matrix")
```

```{r tree-confusion-plot}
autoplot(tree_confusion, type = "heatmap") +
  labs(title = "Decision Tree Confusion Matrix")
```

## ROC curves

```{r roc-curves}
logistic_roc <- roc_curve(
  logistic_predictions,
  truth = high_rating,
  .pred_High,
  event_level = "first"
) %>% mutate(model = "Logistic Regression")

tree_roc <- roc_curve(
  tree_predictions,
  truth = high_rating,
  .pred_High,
  event_level = "first"
) %>% mutate(model = "Decision Tree")

roc_plot <- bind_rows(logistic_roc, tree_roc) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity,
             linetype = model)) +
  geom_path(linewidth = 1, color = "hotpink") +
  geom_abline(linetype = "dotted") +
  coord_equal() +
  labs(
    title = "ROC Curves for IMDb Rating Models",
    x = "False-positive rate",
    y = "True-positive rate",
    linetype = "Model"
  ) +
  theme_minimal()

roc_plot
ggsave("figures/roc_curves.png", roc_plot,
       width = 7, height = 6)
```

# Decision Tree Visualization

```{r tree-visualization}
fitted_tree <- extract_fit_engine(tree_fit)
rpart.plot(
  fitted_tree,
  type = 3,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Decision Tree for Predicting High IMDb Ratings"
)
```

# Cross-Validation

```{r cross-validation}
set.seed(123)
movie_folds <- vfold_cv(movie_train, v = 5,
                        strata = high_rating)
cv_metrics <- metric_set(accuracy, bal_accuracy, roc_auc)

logistic_cv <- fit_resamples(
  logistic_workflow,
  resamples = movie_folds,
  metrics = cv_metrics,
  control = control_resamples(save_pred = TRUE)
)

tree_cv <- fit_resamples(
  tree_workflow,
  resamples = movie_folds,
  metrics = cv_metrics,
  control = control_resamples(save_pred = TRUE)
)

logistic_cv_results <- collect_metrics(logistic_cv) %>%
  mutate(model = "Logistic Regression")

tree_cv_results <- collect_metrics(tree_cv) %>%
  mutate(model = "Decision Tree")

bind_rows(logistic_cv_results, tree_cv_results) %>%
  select(model, .metric, mean, std_err)
```

# Error Analysis

```{r error-analysis}
logistic_errors <- logistic_predictions %>%
  filter(high_rating != .pred_class) %>%
  left_join(
    movie_data %>%
      select(tid, averageRating, runtimeMinutes,
             release_decade, primary_genre, n_regions),
    by = "tid"
  ) %>%
  arrange(desc(.pred_High))

head(logistic_errors, 15)
```

# Results

Complete this section after running the report.

The logistic regression model achieved:

- Accuracy: **0.683**
- Balanced accuracy: **0.532**
- F1 score: **0.235**
- ROC-AUC: **0.483**

The decision tree achieved:

- Accuracy: **0.732**
- Balanced accuracy: **0.566**
- F1 score: **0.267**
- ROC-AUC: **0.611**

Based on **ROC-AUC**, the **decision tree** performed better. Because the dataset is a small sample, the findings should be treated as exploratory.

# Limitations

1. The dataset is a course-provided sample rather than the complete IMDb database.
2. Some movies are missing ratings, runtime, genre, or crew information.
3. The 7.0 cutoff is a project-defined threshold.
4. IMDb ratings can change over time.
5. The model does not include budget, cast popularity, marketing, critic reviews, or release strategy.
6. The models identify associations, not causal relationships.

# Conclusion

This project demonstrates relational joins, data cleaning, feature engineering, visualization, logistic regression, decision trees, cross-validation, model evaluation, and error analysis in R.
